---
title: "Data analysis cycle"
author: 'Utrecht Summerschool: Statistical Programming with R'
date: "2024-06-30"
output: html_document
---

```{r global_opts, include=FALSE}
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE,
                      comment = "",
                      fig.width = 16/2,
                      fig.height = 9/2)

```


```{r Load_packages, echo=FALSE}
# Ensures the package "pacman" is installed
if (!require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse,
               ggplot2,
               gtsummary,
               patchwork,
               GGally,
               ggcorrplot,
               modelsummary,
               kableExtra,
               gt)
```



# Data analysis with mammals data

This is a data set that has information about predation and danger exposure of the mammals during sleep. See here for the data and a description of the variables: https://www.openintro.org/data/index.php?data=mammals


```{r import data}
load("MammalsSleepData.RData")
```

```{r summary statistics}
summary(mammals)
```


## Univariate distributions

```{r univariate distributions}
mammals %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram()
```
```{r bivariate distributions}
mammals %>% 
  pairs(~., data = ., main = "Scatterplot matrix of mammals data")
```

## Bivariate distributions

```{r pairwise other layout}
mammals %>% 
  pairs(~., 
        data = ., 
        col = "blue",       # Change color of the dots
        pch = 19,           # Change shape of points
        cex = 0.5,          # Change size of points
        gap = 0.3,          # Change gaps in between
        main = "Scatterplot Matrix of `mammals` data",
        )

```


## Select subset of the data

Continue analyses with the following variables: `total_sleep`, `brain_wt`, `body_wt`, `life_span`, `danger`.

Remove missing values.

```{r subset data}
subset_mammals <- mammals %>% 
  select(total_sleep, brain_wt, body_wt, life_span, danger) %>%
  na.omit() 

# Alternative:
# subset_mammals <- mammals %>% 
#  select(total_sleep, brain_wt, body_wt, life_span, danger) %>%
#  filter(complete.cases(.))

```

## Summary statistics for subset data

```{r summary statistics subset}
summary(subset_mammals)
```


## Table of summary statistics

```{r table summary stats}
## Nice table with gt_summary

subset_mammals %>% 
  tbl_summary(
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} / {N} ({p}%)")
  ) %>% 
  modify_caption("**Table 1. Summary statistics for subset of mammals data**")
```


## Table summary stats by exposure to danger category

```{r summary stats by danger}
subset_mammals %>% 
  tbl_summary(
    # split by exposure to danger category
    by = danger,
    # change variable labels
    label = list( 
      total_sleep ~ "Total sleep (hours)",
      brain_wt ~ "Brain weight (kg)",
      body_wt ~ "Body weight (kg)",
      life_span ~ "Life span (years)"
      ),
    # change statistics printed in table
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} / {N} ({p}%)")
    ) %>%
  # Change default Characteristic into Variable
  modify_header(label ~ "**Variable**") %>%
  # Add a header text
  modify_caption("**Table 2. Summary statistics by exposure to danger category (1 = less exposure to danger)**") %>% 
  bold_labels()
```

## Explore bivariate relations

```{r plot bivariate relations}
subset_mammals %>% 
  ggpairs(
  lower = list(continuous = wrap("smooth", color = "lightblue")),
  diag = list(continuous = wrap("barDiag", fill = "darkblue", bins=30)),
  upper = list(continuous = wrap("cor", size = 4))
  ) +
  labs(
    title = "Pairwise correlations and distributions of mammals sleep data (subset)",
  ) +
  theme_minimal()
```

## Boxplot of total sleep by exposure to danger

```{r boxlplot total sleep by exposure}
subset_mammals %>%
  ggplot(aes(x = as.factor(danger), y = total_sleep)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Boxplots of total sleep (hours) by exposure to danger",
    x = "Danger category (1=least)",
    y = "Total Sleep (hours)"
  ) +
  theme_minimal()

```
## Lineplot total sleep by life span

```{r lineplot total sleep by life span}
subset_mammals %>% 
  ggplot(aes(x=life_span, y=total_sleep)) + 
  geom_smooth() +
   labs(
    title = "Lineplot total sleep by life_span",
    x = "Life span (years)",
    y = "Total sleep (hours)"
  ) +
  theme_minimal()
```


## Create new variable: brain body ratio

```{r create brain_body variable}
subset_mammals <- subset_mammals %>% 
    dplyr::mutate(brain_body = brain_wt / body_wt) %>% 
    glimpse()
```


## Total sleep and brain body ratio

```{r lineplot sleep and brain body ratio}
subset_mammals %>% 
  ggplot(aes(x=brain_body, y=total_sleep)) + 
  geom_smooth() +
   labs(
    title = "Lineplot total sleep by brain body ratio",
    x = "Ratio brain weight and body weight",
    y = "Total sleep (hours)"
  ) +
  theme_minimal()
```

## Fit linear model

Fit a linear model with dependent variable `total_sleep` and predictors `life_span`, `brain_body` and `danger`.
```{r fil lm mammals subset}
fit_lm_mammals <- lm(total_sleep ~ life_span + brain_body + danger, data = subset_mammals)

summary(fit_lm_mammals)
```


## Regression table with `gtsummary`

Create a table for the output of the linear regression analysis. 

```{r regression table fit lm mammals subset}
## Output of regression analysis with gtsummary()

t1 <-
  fit_lm_mammals %>% 
  tbl_regression(
    intercept = TRUE
  ) %>% 
  modify_header(label ~ "**Predictor**") %>%
  modify_caption("**Table 1. Results of linear regression analysis for total sleep on life span, brain body ratio and exposure to danger**")

t1

```


## Fit linear regression model with interaction between `life_span` and `danger`.

```{r fit lm interaction}
fit_lm_interaction <- lm(total_sleep ~ brain_body + life_span*danger, data = subset_mammals) 
summary(fit_lm_interaction)
```

## Create a table for the regression model with interaction

```{r table lm interaction}
 
t2 <-
  fit_lm_interaction %>% 
  tbl_regression(
    intercept = TRUE
  ) %>% 
  modify_header(label ~ "**Predictor**") %>%
  modify_caption("**Table 2. Results of linear regression analysis for total sleep on brain body ratio and interaction between exposure to danger and life span**")

t2
```

## Test difference between lm models with `anova`

```{r test difference lm models}
anova(fit_lm_mammals, fit_lm_interaction)
```

## Display results of lm models

```{r table lm models}
tbl_merge(
  list(t1, t2),
  tab_spanner = c("**Main effects model**", "**Interaction model**")
) %>% 
  modify_header(label ~ "**Predictor**") %>%
  modify_caption("**Table 3. Results of linear regression analysis for total sleep: main effects model and interaction model**")
```

## Tables with `modelsummary`

https://modelsummary.com/index.html

```{r make table with modelsummary}
models <- list(
  "fit_mammals" = lm(total_sleep ~ life_span + brain_body + danger, data = subset_mammals),
  "fit_mammals_interaction" = fit_lm_interaction <- lm(total_sleep ~ brain_body + life_span*danger, data = subset_mammals) 
)
```

## Make a table with `modelsummary`

```{r modelsummary table lm models}

modelsummary(
  models,
  fmt = 1,
  estimate  = "{estimate} [{conf.low}, {conf.high}]",
  statistic = NULL,
  coef_omit = "Intercept",
  output = "kableExtra")
```

